### 2021-09-14 thesis/tex/sterileResults.tex
update

% \begin{document}
\section{The Sterile Hypothesis}\label{sec:sterileResults}
With the null ($3\nu$) hypothesis normalized to the IceCube Monte Carlo as in Eq.~\ref{eq:Nth}, we are now in good shape to study the sterile effect on the probabilities, and how that compares to data.
First, let us see how the collected data from Fig.~\ref{fig:IC_data} deviates from the predicted $3\nu$ oscillations.
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/IC_rates.pdf}
    \caption{}\label{fig:IC_rates}
\end{figure}
Fig.~\ref{fig:IC_rates} shows an impressive agreement to the standard $3\nu$ oscillation picture, with the largest deviations of $~6\%$
for neutrinos close to double-digit \si{\TeV} values. In zenith, the data only deviates $\pm 1\%$. Recalling the \si{\TeV} disappearance from 
Fig.~\ref{fig:sterile_resonance}, we see that we have a similar deficiency found in the data at \SI{4}{\TeV}. Thus, we expect the best fitting 
sterile hypothesis to include a $\dm[41]$ that places the resonant disappearance in the same region. From Fig.~\ref{fig:resonance_shift}, we see
that a mass of $\dm[41] = \SI{1.5}{\eV \squared}$ achieves this.

\subsection{\texorpdfstring{$\chi^2$}{Chi-squared} Minimization}
For our analyses, we define our $\chi^2$ as
\begin{align} \label{eq:chisq}
    \chi^{2}(\hat{\theta},\alpha,\beta)=\sum_{ij} \frac{\left(N^\text{th}-N^\text{data}\right)_{ij}^{2}}
    {\left(\sigma^\text{data}_{ij}\right)^{2} + \left(\sigma^\text{syst}_{ij}\right)^{2}}+ 
    \frac{(1-\alpha)^2}{\sigma_\alpha^2} + \frac{\beta^2}{\sigma_\beta^2}\,
\end{align}
where we minimize over the model parameters $\hat{\theta}$, the penalty terms $\alpha$ and $\beta$.
$N_{ij}^\text{th}$ is the expected number of events from theory, and $N_{ij}^\text{data}$ is the observed number of events in that bin. 
We set $\sigma_\alpha = 0.25$ as the atmospheric flux normalization error, and $\sigma_\beta = 0.04$ as the zenith angle slope error~\cite{hondapaper}. 
The observed event number has an associated Poissonian uncertainty $\sigma_{ij}^\text{data} = \sqrt{N_{ij}^\text{data}}$.
For IceCube, the event count takes the form
\begin{align}
    N^\text{th}_{ij} = \alpha\left[1+\beta (0.5 + \zreco_i )\right] N_{ij}(\hat{\theta})\,,
\end{align}
with $N_{ij}(\hat{\theta})$ from Eq.~\ref{eq:Nth}. Here, the term $ \beta (0.5 + \zreco_i )$ allows the event distribution to rotate with angle $\beta$ around the median zenith angle of $\zreco = -0.5$.
We also have an uncorrelated systematic error $\sigma_{ijk}^\text{syst}$. We set $\sigma_{ijk}^\text{syst} = f\sqrt{N_{ijk}^\text{data}}$, where $f$ is a percentage of our own choosing.

The minimization of Eq.~\ref{eq:chisq} simply returns a value for each set of model parameters. We use this value to quantify to what extent
our theoretical simulations $N^\text{th}_{ij}$ agree with the data $N^\text{data}_{ij}$ within the error bounds provided by $\sigma_a,\sigma_b,$ and $f$.
We then select the set of $\chi^2$ values which are the smallest and our allowed parameters are then their associated $\hat{\theta}$.
We then translate the $\chi^2$ distribution by subtracting the best-fit point, and analyze $\Delta \chi^2 = \chi^2 - \chi^2_{min}$.
From the $\chi^2$-distribution, one can show that the $90\%$ confidence level has a value of $2.71$ for two degrees of freedom. When slicing through the 
two-dimensional grid of $\Delta \chi^2$ at this level, we then obtain a contour plot that tells us what regions are within our $90 \%$ and what regions are not.

@@

We now in Fig.~\ref{fig:final_rate_plot} plot the best-fit sterile hypothesis with the data from Fig.~\ref{fig:IC_rates}, to finally see how the best 
sterile hypothesis compares to data.
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/final_rate_plot.pdf}
    \caption{The predicted binned event count assuming our best-fit sterile hypothesis, compared with the 
    null and data.}\label{fig:final_rate_plot}
\end{figure}

>>>>

% \begin{document}
\section{The Sterile Hypothesis}\label{sec:sterileResults}
With the null ($3\nu$) hypothesis normalized to the IceCube Monte Carlo as in Eq.~\ref{eq:Nth}, we are now in good shape to study the sterile effect on the probabilities, and how that compares to data.
First, let us see how the collected data from Fig.~\ref{fig:IC_data} deviates from the predicted $3\nu$ oscillations.
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/IC_rates.pdf}
    \caption{How the null hypothesis compares to the IceCube data from~\cite{IC2020}.}\label{fig:IC_rates}
\end{figure}
Fig.~\ref{fig:IC_rates} shows an impressive agreement to the standard $3\nu$ oscillation picture, with the largest deviations of $~6\%$
for neutrinos with high single-digit \si{\TeV} values. In zenith, the data only deviates $\pm 2\%$. Recalling the \si{\TeV} disappearance from 
Fig.~\ref{fig:sterile_resonance}, we see that we have a similar deficiency found in the data at \SI{4}{\TeV}. Thus, we expect the best fitting 
sterile hypothesis to include a $\dm[41]$ that places the resonant disappearance in the same region. From Fig.~\ref{fig:resonance_shift}, we see
that a mass of $\dm[41] = \SI{1.5}{\eV \squared}$ achieves this.

\subsection{\texorpdfstring{$\chi^2$}{Chi-squared} Minimization}
For our analyses, we define our $\chi^2$ as
\begin{align} \label{eq:chisq}
    \chi^{2}(\hat{\theta},\alpha,\beta)=\sum_{ij} \frac{\left(N^\text{th}-N^\text{data}\right)_{ij}^{2}}
    {\left(\sigma^\text{data}_{ij}\right)^{2} + \left(\sigma^\text{syst}_{ij}\right)^{2}}+ 
    \frac{(1-\alpha)^2}{\sigma_\alpha^2} + \frac{\beta^2}{\sigma_\beta^2}\,
\end{align}
where we minimize over the model parameters $\hat{\theta}$, the penalty terms $\alpha$ and $\beta$.
$N_{ij}^\text{th}$ is the expected number of events from theory, and $N_{ij}^\text{data}$ is the observed number of events in that bin. 
We set $\sigma_\alpha = 0.25$ as the atmospheric flux normalization error, and $\sigma_\beta = 0.05$ as the zenith angle slope error~\cite{hondapaper}. 
The observed event number has an associated Poissonian uncertainty $\sigma_{ij}^\text{data} = \sqrt{N_{ij}^\text{data}}$.
For IceCube, the event count takes the form
\begin{align}
    N^\text{th}_{ij} = \alpha\left[1+\beta (0.5 + \zreco_i )\right] N_{ij}(\hat{\theta})\,,
\end{align}
with $N_{ij}(\hat{\theta})$ from Eq.~\ref{eq:Nth}. Here, the term $ \beta (0.5 + \zreco_i )$ allows the event distribution to rotate with angle $\beta$ around the median zenith angle of $\zreco = -0.5$.
We also have an uncorrelated systematic error $\sigma_{ijk}^\text{syst}$. We set $\sigma_{ijk}^\text{syst} = f\sqrt{N_{ijk}^\text{data}}$, where $f$ is a factor of our own choosing.

The minimization of Eq.~\ref{eq:chisq} simply returns a value for each set of model parameters. We use this value to quantify to what extent
our theoretical simulations $N^\text{th}_{ij}$ agree with the data $N^\text{data}_{ij}$ within the error bounds provided by $\sigma_a,\sigma_b,$ and $f$.
We then select the set of $\chi^2$ values which are the smallest and our allowed parameters are then their associated $\hat{\theta}$.
We then translate the $\chi^2$ distribution by subtracting the best-fit point, and analyze $\Delta \chi^2 = \chi^2 - \chi^2_{min}$.
From the $\chi^2$-distribution, one can show that the $90\%$ confidence level has a value of $2.71$ for two degrees of freedom. When slicing through the 
two-dimensional grid of $\Delta \chi^2$ at this level, we then obtain a contour plot that tells us what parameter values are within our $90 \%$ confidence level and what regions are not.

@@

We now in Fig.~\ref{fig:final_rate_plot} plot the best-fit sterile hypothesis with the data from Fig.~\ref{fig:IC_rates}, to finally see how the best 
sterile hypothesis compares to data.
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/final_rate_plot.pdf}
    \caption{The predicted binned event count assuming our best-fit sterile hypothesis, compared with the 
    null and data.}\label{fig:final_rate_plot}
\end{figure}

### 2021-09-14 thesis/tex/sterileTheory.tex
update

%\input{commands}
%\begin{document}
\section{The Sterile State}\label{sec:anomalies}
In 1996, the LSND experiment reported an excess of $\ane$ events from an $\anm$ beam~\cite{lsnd}.  
Nine years later, MiniBooNE not only reproduced the $\ane$ anomaly, but observed an excess in the
$\ne$ events too. Together with the so-called reactor and gallium anomalies, these reports suggested 
that the three massive neutrino framework could be amended.
By introducing a fourth heavy mass state $\nu_4$, both appearance and disappeareance anomalies could be minimally accomodated.
However, we know from the decay width of the $Z$ boson that it only can interact with three flavor species\footnote{
    Direct measurements of the invisible $Z$ decay yields $N=2.92 \pm 0.05$~\cite{pdg}.
}, 
so this fourth mass state cannot be interacting weakly. In other words, it needs to transform as a
singlet under the broken gauge group $SU(2)_L \times U(1)_{EM}$.
We now distinguish between the three original neutrino flavors ($e$, $\mu$, and $\tau$) and the new fourth 
flavor by calling the former \emph{active} neutrinos
and the latter \emph{sterile} due to its non-interacting behavior.

The experiments listed above indicate that the mass-squared difference of the 
sterile neutrino is in the \si{\eV\squared} scale, while the two others
are three and five magnitudes smaller. To remind us of this hierarchy, we write $3+1$, since the sterile state 
proposed is a lot more massive than the active states.

@@

\begin{align}\label{eq:Vsterile}
    V_{4gen} &=
    \begin{pmatrix}
        V_{CC} + V_{NC} & 0 & 0 & 0 \\
        0 &V_{NC} & 0 & 0 \\
        0 & 0 & V_{NC} & 0 \\
        0 & 0 & 0 & 0 
    \end{pmatrix} - V_{NC}\begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 &1 & 0 & 0 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & 1 
    \end{pmatrix}
    = \sqrt{2}G_F\begin{pmatrix}
        N_e & 0 & 0 & 0 \\
        0 &0 & 0 & 0 \\
        0 & 0 & 0 & 0 \\
        0 & 0 & 0 & -N_n/2
    \end{pmatrix} \nonumber \\
    &=\sqrt{2}G_F N_e\begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 &0 & 0 & 0 \\
        0 & 0 & 0 & 0 \\
        0 & 0 & 0 & -1/2
    \end{pmatrix}\,.
\end{align}
where we have assumed electrical neutrality in the last step, yielding $N_e = N_n$.

Thus, the final Hamiltonian with a fourth sterile neutrino is 
\begin{align}\label{eq:H_4gen}
    H_{4gen} &= \frac{1}{2E}\left[U \begin{pmatrix}
            0 & 0 & 0 & 0\\
            0 & \dm_{21} & 0  & 0 \\
            0 & 0 & \dm_{31} & 0 \\
            0 & 0 & 0 & \dm_{41} \\
        \end{pmatrix} U^\dagger\right] + \sqrt{2}G_F N_e 
        \begin{pmatrix}
            1 & 0 & 0 & 0 \\
            0 & 0 & 0 & 0 \\
            0 & 0 & 0 & 0 \\
            0 & 0 & 0 & -1/2
        \end{pmatrix}\,. 
\end{align}

Now, looking at Eqs.~\ref{eq:U4gen_param} and~\ref{eq:dm4gen}, we see that we have introduced four new parameters: $\dm[41]$, $\theta_{14}$
, $\theta_{24}$, $\theta_{34}$. In our analysis, we reduce the number of parameters by only considering $\dm[41]$ and $\theta_{24}$, which are 
the most important for the $\nm$ oscillations. In fact, one can close the $e$ channel by letting $\theta_{14} = 0$ in the \si{\GeV} to \si{\TeV} range, thus 
using a two-flavor neutrino approximation. This is not done in this work.

@@

\begin{figure}
    \centering
    \begin{subfigure}{0.35\textwidth}
        \includegraphics[width=1\linewidth]{figures/sterile_resonance.pdf}
        \caption{$\Pamam$ resonant disappearance}\label{fig:sterile_resonance}
    \end{subfigure}
    \begin{subfigure}{0.69\textwidth}
        \includegraphics[width=1\linewidth]{figures/resonance_shift.pdf}
        \caption{Shifting and amplification of the resonance by altering $\dm[41]$ and $\theta_{24}$, respectively.}\label{fig:resonance_shift}
    \end{subfigure}
\end{figure}

So for a \si{\eV}-scale sterile neutrino and a non-zero $\theta_{24}$, we expect a \si{\TeV} $\anm$ disappearance.
The resonance dip is affected by the value of $\dm[41]$ and $\theta_{24}$ as shown in Fig.~\ref{fig:resonance_shift}.
We see how the value of $\dm[41]$ shifts the peak, while the mixing angle adjusts its strength. 
In the $3+0$ scenario, $\Pamam = 1$ in this range. Thus, if we are able to see a $\anm$ disappearance in the \si{\TeV} region,
we are in a good position to argue for a \si{\eV} scale $\dm[41]$.

>>>>

%\input{commands}
%\begin{document}
\section{The Sterile State}\label{sec:anomalies}
In 1996, the LSND experiment reported an excess of $\ane$ events from an $\anm$ beam~\cite{lsnd}.  
Nine years later, MiniBooNE not only reproduced the $\ane$ anomaly but observed an excess in the
$\ne$ events too. Together with the so-called reactor and gallium anomalies, these reports suggested 
that the three massive neutrino framework could be amended.
By introducing a fourth heavy mass state $\nu_4$, both appearance and disappearance anomalies could be minimally accommodated.
However, we know from the decay width of the $Z$ boson that it only can interact with three flavor species\footnote{
    Direct measurements of the invisible $Z$ decay yields $N=2.92 \pm 0.05$~\cite{pdg}.
}, 
so this fourth mass state cannot be interacting weakly. In other words, it needs to transform as a
singlet under the broken gauge group $SU(2)_L \times U(1)_{EM}$.
We now distinguish between the three original neutrino flavors ($e$, $\mu$, and $\tau$) and the new fourth 
flavor by calling the former \emph{active} neutrinos
and the latter \emph{sterile} due to its non-interacting behavior.

The experiments listed above indicate that the mass-squared difference of the 
sterile neutrino is in the \si{\eV\squared} scale, while the two others
are three and five magnitudes smaller. To remind us of this hierarchy, we write $3+1$, since the sterile state 
proposed is heavier than the active states.

@@

\begin{align}\label{eq:Vsterile}
    V_{4gen} &=
    \begin{pmatrix}
        V_{CC} + V_{NC} & 0 & 0 & 0 \\
        0 &V_{NC} & 0 & 0 \\
        0 & 0 & V_{NC} & 0 \\
        0 & 0 & 0 & 0 
    \end{pmatrix} - V_{NC}\begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 &1 & 0 & 0 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & 1 
    \end{pmatrix} \nonumber \\
    &= \sqrt{2}G_F\begin{pmatrix}
        N_e & 0 & 0 & 0 \\
        0 &0 & 0 & 0 \\
        0 & 0 & 0 & 0 \\
        0 & 0 & 0 & -N_n/2
    \end{pmatrix} \nonumber \\
    &=\sqrt{2}G_F N_e\begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 &0 & 0 & 0 \\
        0 & 0 & 0 & 0 \\
        0 & 0 & 0 & -1/2
    \end{pmatrix}\,.
\end{align}
where we have assumed electrical neutrality in the last step, yielding $N_e = N_n$.
Thus, the final Hamiltonian with a fourth sterile neutrino is 
\begin{align}\label{eq:H_4gen}
    H_{4gen} &= \frac{1}{2E}\left[U \begin{pmatrix}
            0 & 0 & 0 & 0\\
            0 & \dm_{21} & 0  & 0 \\
            0 & 0 & \dm_{31} & 0 \\
            0 & 0 & 0 & \dm_{41} \\
        \end{pmatrix} U^\dagger\right] + \sqrt{2}G_F N_e 
        \begin{pmatrix}
            1 & 0 & 0 & 0 \\
            0 & 0 & 0 & 0 \\
            0 & 0 & 0 & 0 \\
            0 & 0 & 0 & -1/2
        \end{pmatrix}\,. 
\end{align}

Now, looking at Eqs.~\ref{eq:U4gen_param} and~\ref{eq:dm4gen}, we see that we have introduced four new parameters: $\dm[41]$, $\theta_{14}$
, $\theta_{24}$, $\theta_{34}$. In our analysis, we reduce the number of parameters by only considering $\dm[41]$ and $\theta_{24}$, which are 
the most important for the $\nm$ oscillations. In fact, one can supress the $e$ channel by letting $\theta_{14} = 0$ in the \si{\GeV} to \si{\TeV} range, thus 
using a two-flavor neutrino approximation.

@@

\begin{figure}
    \centering
    \begin{subfigure}{0.7\textwidth}
        \includegraphics[width=1\linewidth]{figures/sterile_resonance.pdf}
        \caption{$\Pamam$ resonant disappearance}\label{fig:sterile_resonance}
    \end{subfigure}
    \begin{subfigure}{0.9\textwidth}
        \includegraphics[width=1\linewidth]{figures/resonance_shift.pdf}
        \caption{Shifting and amplification of the resonance by altering $\dm[41]$ and $\theta_{24}$, respectively.}\label{fig:resonance_shift}
    \end{subfigure}
\end{figure}

So for a \si{\eV}-scale sterile neutrino and a non-zero $\theta_{24}$, we expect a \si{\TeV} $\anm$ disappearance.
The resonance dip is affected by the value of $\dm[41]$ and $\theta_{24}$ as shown in Fig.~\ref{fig:resonance_shift}.
We see how the value of $\dm[41]$ shifts the peak, while the mixing angle adjusts its strength. 
In the $3+0$ scenario, $\Pamam = 1$ in this range. Thus, if we are able to see a $\anm$ disappearance in the \si{\TeV} region,
we are in a good position to argue for a \si{\eV^2} scale $\dm[41]$.
